---
title: "Weight Lifting Exercise Prediction"
author: "Tynisha Niles"
date: "2026-07-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Practical Machine Learning Course Project

## Background

Using wearable devices such as accelerometers, it is possible to collect detailed information about physical activity. The goal of this project is to predict the manner in which participants performed a weight lifting exercise. The response variable is **classe**, which has five possible categories (A, B, C, D, and E).

---

# Load Required Libraries

```{r libraries, message=FALSE, warning=FALSE}

library(ggplot2)
library(caret)
library(randomForest)
library(dplyr)
```

---

# Load the Data

The training and testing datasets are imported into R.

```{r load-data}
trainingData <- read.csv(
  "pml-training.csv",
  na.strings = c("NA", "", "#DIV/0!")
)

testingData <- read.csv(
  "pml-testing.csv",
  na.strings = c("NA", "", "#DIV/0!")
)
```

---

# Explore the Data

The dimensions and structure of the datasets are examined before cleaning.

```{r explore-data}
dim(trainingData)
dim(testingData)

str(trainingData)

summary(trainingData)
```

---

# Clean the Data

Columns containing missing values are removed. The first seven variables are also removed because they contain identification information that does not contribute to prediction.

```{r clean-data}
trainingData <- trainingData[, colSums(is.na(trainingData)) == 0]
testingData <- testingData[, colSums(is.na(testingData)) == 0]

trainingData <- trainingData[, -(1:7)]
testingData <- testingData[, -(1:7)]

dim(trainingData)
dim(testingData)

trainingData$classe <- factor(trainingData$classe)
```

---

# Partition the Data

The training dataset is divided into a training set (70%) and a validation set (30%).

```{r partition-data}
set.seed(123)

inTrain <- createDataPartition(
  trainingData$classe,
  p = 0.70,
  list = FALSE
)

trainSet <- trainingData[inTrain, ]
validationSet <- trainingData[-inTrain, ]
```

---

# Cross Validation

Five-fold cross-validation is used while training the model.

```{r cross-validation}
fitControl <- trainControl(
  method = "cv",
  number = 5
)
```

---

# Train the Random Forest Model

A Random Forest model is fitted because it performs well with high-dimensional classification problems.

```{r train-model}
set.seed(123)

rfModel <- train(
  classe ~ .,
  data = trainSet,
  method = "rf",
  trControl = fitControl
)

rfModel
```

---

# Predict the Validation Data

The fitted model is used to predict the validation dataset.

```{r validation-predictions}
validationPredictions <- predict(
  rfModel,
  validationSet
)

head(validationPredictions)
```

---

# Evaluate Model Performance

The confusion matrix is used to evaluate the classification accuracy.

```{r confusion-matrix}
results <- confusionMatrix(
  validationPredictions,
  validationSet$classe
)

results
```

---

# Calculate Accuracy and Out-of-Sample Error

The overall accuracy and estimated out-of-sample error are calculated.

```{r accuracy}
accuracy <- results$overall["Accuracy"]

accuracy

outOfSampleError <- 1 - accuracy

outOfSampleError
```

### Interpretation

The Random Forest model achieved a very high prediction accuracy. The estimated out-of-sample error is the difference between one and the model accuracy. A low out-of-sample error indicates that the model is expected to perform well on unseen data.

---

# Predict the 20 Test Cases

The trained model is applied to the testing dataset.

```{r final-predictions}
finalPredictions <- predict(
  rfModel,
  testingData
)

finalPredictions
```

---

# Save Predictions

The predictions can be exported to a CSV file.

```{r save-predictions}
write.csv(
  finalPredictions,
  "Course_Project_Predictions.csv",
  row.names = FALSE
)
```

---

# Conclusion

A Random Forest classifier was developed using the cleaned accelerometer dataset. Five-fold cross-validation was used during model training to reduce overfitting and estimate prediction performance. The model achieved high accuracy on the validation dataset, resulting in a very low estimated out-of-sample error. Finally, the trained model was used to predict the 20 unseen test cases provided for the course project.